Glyph-maps for visualizing climate data and models

### 2011-09-23 paper.tex
finished scale section

*** Global/local min/max, remove intercept/average, data vs model scale
*** This section needs to have plots re-done, once we have decided on the key pieces

Scaling of the data can produce dramatically different emphases on
structure. For spatiotemporal data the most natural scale is to use
the global minumum and maximum. The resulting icons will have size
relative to the magnitude of the full range of the data. Locations
with overall large or small temperature will be the focus. Figure
\ref{fig:glyph-map} shows data scaled to -1 to 1 based on the global
minimum and maximum, using:

\begin{equation}
y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} \times 2 - 1
\label{scale1}
\end{equation}

\noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points. *** Reference other plots that use global scale too.

Alternatively, scaling locally at each spatial location enables
exploring the seasonal scale, or shape differences, from one location
to another. Fig \ref{***} uses local scaling using:

\begin{equation}
y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} \times 2 - 1
\label{scale2}
\end{equation}

\noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$. With the local scaling, the shape of the icon, for example, differences in the seasonal patterns over spatial locations, are emphasized.
Figure \ref{***} shows data that has been locally scaled. We can see *** The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}.

It can also be advantageous to relocate the data uniformly, by extracting the average at each location. Particularly for polar glyphs the average value can lead to some unintuitive patterns in the glyphs. In this situation we would use $y_{st}-\bar{y}_s$ instead of $y_{st}$ in equations \ref{scale1}, \ref{scale2}, with minima and maxima calculated on the mean-shifted data.

Another way to think about these scale options is partitioning the data structure into shape, magnitude and shift components. The different ways to scale emphasize these different components. The shift component can be extracted from the data in different ways, using the intercept from a linear model, for example, instead of the average.

**** This is different - I think it just confuses the topic of scaling. Because the GAM is not used to rescale the data, you still have to do that using the local or global scling equations above. Moving the discussion to ``What to plot''.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.33\linewidth]{month-rescale-none}%
  \includegraphics[width=0.33\linewidth]{month-rescale-max}%
  \includegraphics[width=0.33\linewidth]{month-rescale01}
  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) unscaled, (center) scaled to have maximum 1, (right) scaled to have maximum 1 and minimum 0.}
  \label{fig:scaling}
\end{figure}

A possible compromise is to use line colour to display the standardising variable, as shown in Figure~\ref{fig:scaling-col}.  This makes it possible to see the individual patterns while still retaining some ability to see which shapes represent small absolute changes.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.33\linewidth]{month-rescale01-col}
  \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.}
  \label{fig:scaling-col}
\end{figure}

Other types of scaling may be interesting to explore, such as a classical scaling using the mean and standard deviation.

Equivalent to using individual scales in facetted graphics.

@@

*** Connection to small multiples, facetted plots

>>>>

%*** Global/local min/max, remove intercept/average, data vs model scale
%*** This section needs to have plots re-done, once we have decided on the key pieces

Scaling of the data, into the values used for each glyph, emphasizes different aspects of the data. For spatiotemporal data the most natural scale is to use the global minumum and maximum, because the temporal variables are measuring the same information, and we want to make comparisons across the spatial region. The icons will have size relative to the magnitude of the full range of the data. Locations with overall large temperature will catch the focus. Figure \ref{fig:nasa-deseas-glyph} shows data scaled to 0 to 1 based on the global minimum and maximum, using:

\begin{equation}
y_{st} = \frac{y_{st}-y_{min}}{y_{max}-y_{min}} %\times 2 - 1
\label{scale1}
\end{equation}

\noindent where $y$ is the variable used on the minor axis in equations \ref{coords.eqn}, \ref{coords.polar.eqn}, $y_{min} = \mbox{minimum}\{y_{st}; s=1, \dots S, t=1, ..., T \}, y_{max} = \mbox{maximum}\{y_{st}; s=1, \dots S, t=1, ..., T\}$, $S=$ is the number of spatial locations and $T=$ is the number of time points. In addition, $y_{st}$ may be further scaled by doubling and subtracting 1 to give values between -0.5 and 0.5 for using in equations \ref{coords.eqn}, \ref{coords.polar.eqn}.

Alternatively, scaling locally at each spatial location enables exploring shape differences, particularly seasonal patterns, from one location to another:

\begin{equation}
y_{st} = \frac{y_{st}-y_{s,min}}{y_{s,max}-y_{s,min}} %\times 2 - 1
\label{scale2}
\end{equation}

\noindent where $y_{s, min} = \mbox{minimum}\{y_{st}; t=1, ..., T \}, y_{s,max} = \mbox{maximum}\{y_{st}; t=1, ..., T\}$. With the local scaling, the shape of the icon, for example, differences in the seasonal patterns over spatial locations, are emphasized.  Fig \ref{fig:scaling} shows glyph-maps generated usng both global scaling (left) and local scaling (right) for the NASA temperature data. On the global scale, the El Ni\~no blip in temperature is primarily visible, along with the generally flat series over most water locations, and somewhat increasing or variable temerpatures over land. The local scale emhpasizes the individual shapes: The El Ni\~no pattern is visible, but extend to a larger spatial region including the Caribbean, dramatic linear increases across the Andes, decreases in the Gulf of Mexico. (The local scale is equivalent to using individual scales in the facetted graphics of \texttt{ggplot2}). One caution, with local scale, the viewer needs to realize that the scale is not relative, big patterns in some locations might be just tiny effects. To indicate this on the locally scaled data plots, using a very pale coloring to code the original scale in the tiles, might help puts patterns in perspective.

Other types of shifting and scaling can be useful. The mean and standard deviation might be used to standardize values, or the median and interquartile range, might be useful with some types of data. Shifting the values at each location to have mean zero, places more focus on the trend. This is particularly important for the star glyphs, where differences in the average value lead to unintuitive patterns in the resulting glyphs.  In this situation we would use $y_{st}-\bar{y}_s$ instead of $y_{st}$ in equations \ref{scale1}, \ref{scale2}, with minima and maxima calculated on the mean-shifted data.

%Another way to think about these scale options is partitioning the data structure into shape, magnitude and shift components. The different ways to scale emphasize these different components. The shift component can be extracted from the data in different ways, using the intercept from a linear model, for example, instead of the average.

%**** This is different - I think it just confuses the topic of scaling. Because the GAM is not used to rescale the data, you still have to do that using the local or global scling equations above. Moving the discussion to ``What to plot''.

\begin{figure}[htbp]
  \centering
  %\includegraphics[width=0.33\linewidth]{month-rescale-none}%
  \includegraphics[width=0.5\linewidth]{month-rescale-max}%
  \includegraphics[width=0.5\linewidth]{month-rescale01}
  \caption{Smoothed de-seasonalized daily temperature glyph-map. (Left) %unscaled, (center) 
globally scaled, (right) locally scaled.}
  \label{fig:scaling}
\end{figure}

%A possible compromise is to use line colour to display the standardising variable, as shown in Figure~\ref{fig:scaling-col}.  This makes it possible to see the individual patterns while still retaining some ability to see which shapes represent small absolute changes.

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=0.33\linewidth]{month-rescale01-col}
%  \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions.}
%  \label{fig:scaling-col}
%\end{figure}

%Other types of scaling may be interesting to explore, such as a classical scaling using the mean and standard deviation.

%Equivalent to using individual scales in facetted graphics.

One more important point: if a model summary is displayed in the glyph-map, such as the predicted values, it may be important to force the scale to that of the original data. Displaying the predicted values in the absence of the raw data scale can skew the interpretation of the magnitude of the effect.

@@
